Historical Soil Organic Carbon Budget

### 2020-06-18 SOCbudgetPaper/discussion.Rmd
Update on method (tillage, irrigatoin, initialization, some smaller comments) + Update intro

>>>>

- pasture dynamics are neglected and treated as natveg

- no tillage adaptaion is neglected on cropland. Pastures are assumed to be not tilled at all (propably at least heavy managed pastures are tilled with some rotation)

- irrigated areas are no crop specific and growing periods are not considered (since this is very complex to calculate average growing periods). Crop specific growing periods might be possible using LPJmL data.

### 2020-06-18 SOCbudgetPaper/introduction.Rmd
Update on method (tillage, irrigatoin, initialization, some smaller comments) + Update intro

Introduction text goes here.
You can change the name of the section if neccessary using `\introduction[modified heading]`.

>>>>

<!-- Relevance of the general topic
- why important (rescue the world)
Stand der Forschung, sher spezifisch auf deine fragestellung
- Gibt viel biophysikalische Modellierung
Forschungslücke (fängt mit "However" an.)
- Zu wenig Fokus auf management, obwohl sehr einflussreich
Fragestellung, Subfragestellungen
Struktur der Arbeit
-->

<!-- You can change the name of the section if neccessary using `\introduction[modified heading]`. -->

Within the last centuries the recognition of soil organic carbon (SOC) has developed from being a major source of agricultural carbon emissions towards a potentially large additional sink to offset other green house gas emissions and thus mitigate climate change. Whereas the extent of sink potential by SOC enhancement is highly debated (add refs), it is largely agreed upon that the SOC pool itself is the biggest terrestrial carbon pool, over topping the atmospheric and even more the biospheric carbon pool multiple times. Even small changes in SOC drivers might lead to substantial shifts in earth carbon cycled and influence the atmospheric CO2 concentration (ref. permafrost melting). The specific amount of carbon stored in the soil is uncertain though and estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Bathjes, 1996).

Mapping the worlds SOC has researches intensivly over the last decades and led to an increased quality of SOC maps as well as a even better understanding of factors driving magnitude, distribution and dynamics of SOC pools. It is undisputed that natural properties like climatic, biophysical and landscape characteristics play the most important role in this regard. Nevertheless,
these factors can not be easily altered by human intervention, leaving land cover, land use and land management changes the most important factors for SOC dynamics till dawn of civilization <!--to heroic find another word-->.

Recent studies identified the SOC debt of anthropogenic source at around 116 GtC (Sanderman et al.), which compares to older estimates of around 60-130 GtC <!--look up --> (Lal, 2006). Other studies have focus more closely on spatially disaggregation SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemcial processes within SOC dynamics  (<!-- find microbial study -->). Most of these studies doing cutting edge research in their field, but seem to lack proper data on land management.

On the other hand small-scale models (ref. Daycent, RothC, Ecosse, C-Tool) are able to represent well SOC dynamics on field-scale driven by detailed management data on yield levels, fertilizer inputs and various other on farming activities. Due to the lack of proper, concise management data to feed these input demanding models on larger extends, it seems still very complex to scale them up to a global extent.

This study however wants to contribute filling the gap on impacts of agricultural management like yield levels, residue recycling, manure amendments, irrigation and tillage on SOC stocks of the worlds croplands. It is doing so by combining work on agricultural management data on global level with a robust but light weighted SOC model to estimate SOC stocks and stock change factors as well as organic carbon flow dynamics within the agricultural system.

This is critically important since managing agricultural soil might be one of the few setscrew for humans to _naturally_ combat climate change (ref. to natural climate solutions). Other _natural_ options, which might be summarized by restoring potential natural stocks via renaturation of managed land, seem to be often limited due to increased pressure on the land systems for feed, food and energy demand of a growing and developing population on earth.

This paper will introduce in the the first part of the method section the basic concept of SOC dynamics. We continue in the second part with a detailed description of the used global gridded management data on production levels, residue recycling rates, manure amendments, irrigation and tillage adoption. In a third, small part, we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines. 
In our result section we will focus on the SOC dynamics on global croplands by analysing (1) the spatial explicit SOC distribution and depletion over the last decades, (2) compare the climate zone specfic stock changes with default stock changes factors from the IPCC Tier 1 approach and (3) compare global agricultrual carbon flows and stocks to contextualize the importants of the different management aspects.

Finally, we will discuss our findings and their implications on SOC model development and conlude with an analysis and outlook on the the opportunities for SOC management for climate change mitigation and negative emissions.

<!--
* sanderman
* WISE
* SoilGrids
* Stoorvogel
* HANPP

* Smith/Powlson
* Lal
* Bathjes
-->

### 2020-06-18 SOCbudgetPaper/methods.Rmd
Update on method (tillage, irrigatoin, initialization, some smaller comments) + Update intro

BB: I would include the t, also to show that the steady-state is time-dependent.

@@

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types }
 \begin{tabular}{l l l l}
 \tophline
  land-use types BB: table header should be bold   & source of carbon inputs & data source & nitrogen and lignin content \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}
```

@@

This equation can also be reformulated to a massbalance equation as follows:

\begin{equation}
SOC_{t} = SOC_{t-1} - k_{t} \cdot 1\unit{a} + C^{\textrm{in}}  \qquad \forall lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

To initialize all SOC sub-pools we assume that cropped land 
natural vegetation 
* steady-states or 
* spin up

@@

Agricultural management data is estimated using the R library moinput ([cite git link]), which compiles country-specific FAO production and cropland statistics ([@FAOSTAT]) to a comprehensive and constistent data suite. BB: Isnt the whole calculation based on R libraries?
The data is prepared in 5 year time steps from 1965 to 2010, which also restricts our analysis to this time span.
For all the following data, if not declared differently, we interpolate values linearly between the time steps and hold it constant before the first time step from 1961 to 1965. BB: too short for a spin-up, so dont use this word. Maybe also just ignore this detail.

@@

#### Crop Residue Production
Crop residue production and management is based on a revised methodology of ([@bodirsky2012]) and will be explained in key aspects again due to its central role for soil carbon modelling. Starting from crop production estimates of the harvested organs and their respective crop area, we estimate above-ground (ag) and below-ground (bg) residual biomass using yield-dependent <!-- do not know what was meant with this anymore (why "yield-dependent"?). I guess that has to be explained again than -.- --> harvest indices and shoot:root ratios. We assume that all bg residues are recycled to the soil, whereas ag residues can be burned or harvested for other purposes such as feeding animals ([@weindl]), fuel or for material use.
    BB: One new thing is here that we use a doublecropping factor to account for lower harvest indices of multicropping vs a very high single-cropped yield

@@

#### Pasture Production
Using livestock production statistics as well as feed mix assumptions as describted in ([@weindl]) we estimating country specific pasture production. Following the same approach as for crop production we disaggregate and calibrate half-degree pasture production pattern from grass yields from LPJmL and pasture area and rangeland patterns ( (see \@ref(#sec:landuse))) to derive half-degree pasture production patterns.
BB: Oh, does that mean we assume that only the pasture grows that is being eaten? Thats not too much, is it? Maybe ask Isabelle for a better solution (e.g. direct pasture yields from LPJmL). And then the "pasture production" is just the part being removed by grazing. Does the IPCC specifiy something in respect to pastures?

@@

BB: What does not become clear to me: Do the extensive cows only eat pasture, and the intensive one only crops? Does that in the end mean we just redistirbute grazed pasture to pastures and crop-based feed to croplands proportional to production? Or is there more to it?

@@

### Irrigation (100)
BB: Where does the irrigated area come from?
Simple growing period calculations together with irrigation shares of LUH2v2 are used to estimate water effects on decay rates.

### Tillage (100)
Tillage data sets of [Vera, others] together with rules are used to drive tillage effect on decay rates.

>>>>

<!-- BB: I would include the t, also to show that the steady-state is time-dependent. -->

@@

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types }
 \begin{tabular}{l l l l}
 \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}
```

@@

Reformulated equation \ref(eq:steadystate), we obtain a massbalance equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation \ref(eq:inoutflow))}}  \qquad \forall lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

To initialize all SOC sub-pools we assume that cropped land as well as natural vegetation are in a steady state for the specific configuration present within the initilzation year 1961. We assume after a spin-up period of 15 years reliable results start from 1975 and improve over time, since dependency for the initial conditions will decrease.

@@

Agricultural management data is estimated using the R library mrcommons ([cite git link]), which compiles country-specific FAO production and cropland statistics ([@FAOSTAT]) to a comprehensive and constistent data suite. <!-- BB: Isnt the whole calculation based on R libraries? - KK: Sure but I thought makig the relevenat packages visible here makes sense? --> 
The data is prepared in 5 year time steps from 1965 to 2010, which also restricts our analysis to this time span.
For all the following data, if not declared differently, we interpolate values linearly between the time steps.

@@

#### Crop Residue Production
Crop residue production and management is based on a revised methodology of ([@bodirsky2012]) and will be explained in key aspects again due to its central role for soil carbon modelling. Starting from crop production estimates of the harvested organs and their respective crop area, we estimate above-ground (ag) and below-ground (bg) residual biomass using yield-dependent <!-- do not know what was meant with this anymore (why "yield-dependent"?). I guess that has to be explained again than -.- --> harvest indices and shoot:root ratios. We assume that all bg residues are recycled to the soil, whereas ag residues can be burned or harvested for other purposes such as feeding animals ([@weindl]), fuel or for material use.
    <!-- BB: One new thing is here that we use a doublecropping factor to account for lower harvest indices of multicropping vs a very high single-cropped yield - KK: ASk Benni again about this -->

@@

#### Pasture Production
Using livestock production statistics as well as feed mix assumptions as describted in ([@weindl]) we estimating country specific pasture production. Following the same approach as for crop production we disaggregate and calibrate half-degree pasture production pattern from grass yields from LPJmL and pasture area and rangeland patterns ( (see \@ref(#sec:landuse))) to derive half-degree pasture production patterns.
<!-- BB: Oh, does that mean we assume that only the pasture grows that is being eaten? Thats not too much, is it? Maybe ask Isabelle for a better solution (e.g. direct pasture yields from LPJmL). And then the "pasture production" is just the part being removed by grazing. Does the IPCC specifiy something in respect to pastures?  - KK: Clarify with Benni.But actually pastrue production does not matter since just the patterns are used to distribute livestock. Anyway are word on the neglected SOC cycling on pasture compared to natveg has to come somewhere. -->

@@

<!-- BB: What does not become clear to me: Do the extensive cows only eat pasture, and the intensive one only crops? Does that in the end mean we just redistirbute grazed pasture to pastures and crop-based feed to croplands proportional to production? Or is there more to it? - KK: Now acctually not.-->

@@

### Irrigation (100)

We use irrigation area shares to modify the water effect on decomposition by weighting the irrigated and rainfed water factors based on these shares. The LUH2v2 data set provides irrigated fractions for their cropland subcategories. We apply aggregated area shares, leading to the same impact of irrigation of all our crop groups.   
Moreover we assume the irrigation effect to te present for all month of a year, that has been marked as irrigated.

### Tillage (100)

To account for tillage distribution between the three tillage types specfied by the IPCC - full tillage, reduced tillage and no tillage -, we use the simple assumptions that all natural land and pastures are not tilled, whereas annual crops are under full and perennials under reduced tillage. This rules are inspired by the approach of [@porvolik], but neglects the less common adoption of no tillage farming on agricultural land.

Tillage data sets of [Vera, others] together with rules are used to drive tillage effect on decay rates.
